### 2014-10-20 TransitionsUseful.tex
more random changes. now figures are correct
small edits
Pass up to Section 4

For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s, \ell, s') \in
A$, and we write $A_0$ for $A$'s initial state. The intuition is that $A$ is
one process within a concurrent system. The label $\ell$ from set $\mathcal L$
of labels is produced when $A$ participates in some IO exchange with another
process; otherwise it is an empty or ``silent'' label $\epsilon$. As a
shorthand, we sometimes omit labels for $\epsilon$ steps.

@@

\begin{defn}\label{refines}
For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
be a function that is able to replace labels with alternative
labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
sets $\mathcal L$. We say that
\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
\begin{multline*}
\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
(s_A = H \Leftrightarrow s_B = H)
\end{multline*}
Each label in the trace is replaced by the mapping of $f$ on it, and
labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
to denote that multi-label version when applied to $\eta$.
\end{defn}

@@

\begin{figure}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
%\centering
\inference
[Halt]
{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
{\io{SC}{\theta,m}{H}{}}

@@

\section{Decomposing a multi-processor system}
\label{sec:store-atomicity}

@@

Any multi-processor component can be divided into 3 components as shown in
Figure \ref{zoom}. 
We show the top-level system design at top right, and we zoom in on boxes standing for the
memory system and a single processor ($P_i$).
The processor component $P_i$ can be implemented in a
variety of ways, from a simple in-order issue, to a complex out-of-order issue
core, with aggressive speculation. The memory component is normally implemented
using a hierarchy of caches, in order to increase the performance of the
overall system, since the latency of accessing a memory directly is very slow
compared to accessing a much smaller cache. The component between each
processor and the global memory sub-system contains local buffers, $B_i$, specific to
the processor $P_i$ they are connected to.

\begin{figure}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Load]
{}
{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a))}}

@@

Independent of the memory models, the memory component always implements the
\emph{store atomicity} property. This property can be understood using the
behavior defined by the operational semantics of Figure \ref{$M_m$}. The
different behaviors permitted by different memory models arise merely because
of the (1) the reordering in the issue of memory instructions by the core $P_i$
if it can issue instructions out-of-order and (2) the properties of the local
buffers $B_i$ connected to each processor $P_i$. For instance, if $B_i$ was a
store buffer, into which the stores from an in-order processor $P_i$ that are
``committed'' are enqueued, and if any ``subsequent'' load from the same
processor $P_i$ reads the latest value from the store buffer, then one can
establish that such a multi-processor system implements the TSO
\cite{x86tsocacm10} memory model.

The memory component is composed of a \emph{hierarchy of caches}, where each
cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
do our best to satisfy all memory requests, to avoid the latency of a
round-trip with main memory.  With an L1 cache miss, we may still find a hit in
a parent cache below the level of main memory, realizing a smaller but still
significant speedup.  Therefore it is the responsibility of the hierarchy of
caches, which form the memory sub-component, to implement the store atomicity
property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
cache coherence protocol is to establish this invariant for the memory
sub-component.  Concretely, we have verified a \emph{directory-based} protocol
for coordinating an arbitrary tree of caches, where each node stores a
conservative approximation of its children's states.

As an instance of the above decomposition, we will prove that a multi-processor
system with no local buffering in between the processor and the memory
components indeed implements sequential consistency. We implement a highly
speculative processor which executes instructions and issues loads out of
order, but commits instructions (once some ``verification'' is done) in order.

The processor itself can be decomposed into several components. In the zoomed
in version of Figure \ref{zoom}, we show a highly speculative, out-of-order
issue processor. We have the normal architectural state, such as values of
registers.  Our proofs are generic over \emph{a family of instruction set
architectures}, with parameters for opcode sets and functions for executing
opcodes and decoding them from memory.  Other key components are a \emph{branch
predictor}, which guesses at the control-flow path that a processor will
follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
decides which instructions along that path to try executing ahead of schedule.
Our proofs apply to an arbitrary branch predictor, and they work for any
reorder buffer satisfying a simple semantic condition.

Figure~\ref{proofs} gives the overall proof structure that we employ to verify
the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
represents the speculative out-of-order processor, $M_c$ the cache memory,
$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
instruction atomically thus implementing sequential consistency. Notation
$A^n$ is for $n$ copies of system $A$ running in parallel.
$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
one system implements another.  We go into more detail on the symbols from
the figure in the rest of the paper.

@@

\begin{figure}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Halt]
{\dec(s,\pc) = H}
{\io{P$_\text{ref}$}{s,\pc}{H}{}}

@@

\begin{figure*}[t]
\centering
\begin{boxedminipage}[c]{.95\textwidth}
\inference[Fetch]
{}
{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}

@@

\begin{figure*}[t]
\begin{inv}
If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}

@@

\begin{figure*}
\centering

>>>>

For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
intuition is that $A$ is one process within a concurrent system. The
label $\ell$ from set $\mathcal L$ of labels is produced when $A$
participates in some IO exchange with another process; otherwise it is
an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
omit labels for $\epsilon$ steps.  Note that the IO exchange can be
input, output, or both, and its interpretation is dependent on the
context.

@@

\begin{defn}\label{refines}
For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
be a function that is able to replace labels with alternative
labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
set $\mathcal L$. We say that
\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
\begin{multline*}
\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
(s_A = H \Leftrightarrow s_B = H)
\end{multline*}
Each label in the trace is replaced by the mapping of $f$ on it, and
labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
to denote that multi-label version when applied to $\eta$.
\end{defn}

@@

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
%\centering
\inference
[Halt]
{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
{\io{SC}{\theta,m}{H}{}}

@@

\section{Decomposing a Multi-Processor System}
\label{sec:store-atomicity}

@@

Most any conventional multi-processor system can be divided into 3 components as shown in
Figure \ref{zoom}. 
We show the top-level system design at top right, and we zoom in on boxes standing for the
memory system and a single processor ($P_i$).
The processor component $P_i$ can be implemented in a
variety of ways, from literal execution of instructions in program order, to complex speculation
with instructions run simultaneously to exploit parallelism. The memory component is normally implemented
using a hierarchy of caches, in order to increase the performance of the
overall system, since the latency of accessing a memory directly is very slow
compared to accessing a much smaller cache. The component between each
processor and the global memory sub-system contains local buffers, $B_i$, specific to
the processor $P_i$ they are connected to.

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Load]
{}
{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a))}}

@@

Popular instruction set architectures do not guarantee sequential
consistency.  That is, their behavior makes it unsound to imagine that
the different processors of a shared-memory system are running
according to simple interleaving, where each instruction executes
atomically.  However, we want to emphasize that, in every weak-memory
system we are aware of, \emph{the main memory still exposes atomic
  loads and stores}!  The weaker semantics arises only because of (1)
reordering of memory instructions by the core $P_i$ and/or (2) the
properties of the local buffers $B_i$ connected to each processor
$P_i$.  For instance, if $B_i$ were a store buffer, into which the
committed stores from an in-order processor $P_i$ are enqueued, and if
any ``subsequent'' load from the same processor $P_i$ reads the latest
value from the store buffer, then one can establish that such a
multi-processor system implements the TSO~\cite{x86tsocacm10} memory
model.

So, we focus on this opportunity to simplify proof decomposition.  We
will prove that our main memory component satisfies an intuitive
\emph{store atomicity} property.  Despite the potential red herring of
the word ``atomicity,'' this specification is appropriate even for
implementation of weaker memory models than SC.  Store atomicity can
be understood via the operational semantics of Figure \ref{$M_m$},
describing an LTS that receives load and store requests ($\ld$ and
$\st$) from processors and sends back load responses ($\ld\resp$).

The memory component is composed of a \emph{hierarchy of caches}, where each
cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
do our best to satisfy all memory requests, to avoid the latency of a
round-trip with main memory.  With an L1 cache miss, we may still find a hit in
a parent cache below the level of main memory, realizing a smaller but still
significant speedup.  Therefore it is the responsibility of the hierarchy of
caches, which forms the memory sub-component, to implement the store atomicity
property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
cache-coherence protocol is to establish this invariant for the memory
sub-component.  Concretely, we have verified a \emph{directory-based} protocol
for coordinating an arbitrary tree of caches, where each node stores a
conservative approximation of its children's states.

As an instance of the above decomposition, we will prove that a multi-processor
system with no local buffering in between the processor and the memory
components indeed implements sequential consistency. We implement a highly
speculative processor that executes instructions and issues loads out of
order, but commits instructions (once some ``verification'' is done) in order.

The processor itself can be decomposed into several components. In the zoomed-in
version of Figure \ref{zoom}, we show a highly speculative, out-of-order
issue processor. We have the normal architectural state, such as values of
registers.  Our proofs are generic over \emph{a family of instruction set
architectures}, with parameters for opcode sets and functions for executing
opcodes and decoding them from memory.  Other key components are a \emph{branch
predictor}, which guesses at the control-flow path that a processor will
follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
decides which instructions along that path to try executing ahead of schedule.
Our proofs apply to an arbitrary branch predictor, and they work for any
reorder buffer satisfying a simple semantic condition.

Figure~\ref{proofs} gives the overall proof structure that we employ to verify
the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
represents the speculative out-of-order processor, $M_c$ the cache memory,
$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
instruction atomically, thus implementing sequential consistency.

@@

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Halt]
{\dec(s,\pc) = H}
{\io{P$_\text{ref}$}{s,\pc}{H}{}}

@@

\begin{figure*}[t]
\small
\centering
\begin{boxedminipage}[c]{.95\textwidth}
\inference[Fetch]
{}
{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}

@@

\begin{figure*}[t]
\small
\begin{inv}
If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}

@@

\begin{figure*}
\small
\centering
